---
title: "Example R Analysis (replace with your project)"
description: "One-sentence summary of the problem and what you found."
date: "2026-01-20"
categories: [R, Statistics]
format:
  html:
    code-fold: true
---

## Overview

This is how a plain `.R` script becomes a page on the site. Quarto renders
`.qmd` files (R Markdown / Quarto Markdown) using R code chunks — each chunk
runs and its output (tables, plots, printed results) is embedded in the page.

Two ways to bring in your real R work:

1. **Fastest:** run `scripts/convert_r_to_qmd.py path/to/your-script.R` (included
   in this repo) — it wraps your `.R` file into a `.qmd` like this one automatically,
   with a title/description/date placeholder for you to fill in.
2. **Manual:** copy your R code into the chunk(s) below by hand and add the
   YAML header shown above (title, description, date, categories) — that
   header is what powers the Projects listing cards.

## Data

```{r}
#| label: setup
library(tidyverse)

df <- tibble(
  x = 1:10,
  y = (1:10)^2
)
```

## Analysis

```{r}
#| label: sample-plot
#| fig-cap: "Sample output (delete me)"
ggplot(df, aes(x, y)) +
  geom_line() +
  geom_point() +
  labs(title = "Replace this with your real analysis") +
  theme_minimal()
```

```{r}
summary(df)
```

## Notes

- If your real script reads a data file, commit that file to the repo (e.g.
  a `data/` folder) and reference it with a relative path.
- If a script needs an R package beyond the tidyverse, add it to the
  `packages:` list in `.github/workflows/publish.yml` under "Install R
  dependencies".
